---
title: "AAT - Double Difference of Medians"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{aat_double_diff_of_medians}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(splithalfr)
```
This vignette describes a scoring method similar to [Heuer, Rinck, and Becker (2007)](https://doi.org/10.1016/j.brat.2007.08.010); double difference of median reaction times (RTs) for correct responses on Approach Avoidance Task data. It is a subtraction comparing approach bias towards test stimuli relative to approach bias towards control stimuli (avoid_test - approach_test) - (avoid_control - approach_control).

<br />

# Dataset
Load the included AAT dataset and inspect its documentation.
```
data("ds_aat", package = "splithalfr")
?ds_aat
```

## Relevant variables
The columns used in this example are:

* UserID, which identifies participants
* block_type, in order to select only assessment blocks
* appr, approach or avoid trial
* cat, in order to compare test and control stimuli
* response, in order to select only correct first responses
* rt, in order to calculate medians for avoid_test, approach_test, avoid_control, and approach_control
* stim, stimulus ID

## Preprocessing
Only select trials from assessment blocks. 
```
ds_aat <- subset(ds_aat, block_type == "assess")
```

## Counterbalancing
The variables `appr` and `stim` were counterbalanced. Below we illustrate this for the first participant.
```
ds_1 <- subset(ds_aat, UserID == 1)
table(ds_1$appr, ds_1$stim)
```

<br />


# Scoring the AAT

## Scoring function

The scoring function calculates the score of a single participant as follows:

1. selects only correct responses
2. calculates the median RT of remaining responses

```
fn_score <- function (ds) {
  median_avoid_test <- median(
    ds[ds$appr == "no" & ds$cat == "test" & ds$response == 1, ]$rt
  )
  median_approach_test <- median(
    ds[ds$appr == "yes" & ds$cat == "test" & ds$response == 1, ]$rt
  )
  median_avoid_control <- median(
    ds[ds$appr == "no" & ds$cat == "control" & ds$response == 1, ]$rt
  )
  median_approach_control <- median(
    ds[ds$appr == "yes" & ds$cat == "control" & ds$response == 1, ]$rt
  )
  return (
    (median_avoid_test - median_approach_test) - 
    (median_avoid_control - median_approach_control)
  )
}
```    

## Scoring a single participant
Let's calculate the AAT score for the participant with UserID 14. NB - This score has also been calculated manually via Excel in the splithalfr repository.
```
fn_score(subset(ds_aat, UserID == 14))
```

## Scoring all participants
To calculate the AAT score for each participant, we will use R's native `by` function and convert the result to a data frame.
```
scores <- by(
  ds_aat,
  ds_aat$UserID,
  fn_score
)
data.frame(
  UserID = names(scores),
  score = as.vector(scores)
)
```

<br />

# Estimating split-half reliability

## Calculating split scores
To calculate split-half scores for each participant, use the function `by_split`. The first three arguments of this function are the same as for `by`. An additional set of arguments allow you to specify how to split the data and how often. In this vignette we will calculate scores of 1000 permutated splits. The trial properties `app` and `stim` were counterbalanced in the AAT design. We will stratify splits by these trial properties. See the vignette on splitting methods for more ways to split the data.

The `by_split` function returns a data frame with the following columns:

* `participant`, which identifies participants
* `replication`, which counts replications
* `score_1` and `score_2`, which are the scores calculated for each of the split datasets

*Calculating the split scores may take a while. By default, `by_split` uses all available CPU cores, but no progress bar is displayed. Setting `ncores = 1` will display a progress bar, but processing will be slower.*

```
split_scores <- by_split(
  ds_aat,
  ds_aat$UserID,
  fn_score,
  replications = 1000,
  stratification = paste(ds_aat$app, ds_aat$stim)
)
```

## Calculating reliability coefficients
Next, the output of `by_split` can be analyzed in order to estimate reliability. By default, functions are provided that calculate Spearman-Brown adjusted Pearson correlations (`spearman_brown`), Flanagan-Rulon (`flanagan_rulon`), Angoff-Feldt (`angoff_feldt`), and Intraclass Correlation (`short_icc`) coefficients. Each of these coefficient functions can be used with `split_coef` to calculate the corresponding coefficients per split, which can then be plotted or averaged via a simple `mean`. A bias-corrected and accelerated bootstrap confidence interval can be calculated via `split_ci`.

```
# Spearman-Brown adjusted Pearson correlations per replication
coefs <- split_coefs(split_scores, spearman_brown)
# Distribution of coefficients
hist(coefs)
# Mean of coefficients
mean(coefs)
# Confidence interval of coefficients
split_ci(split_scores, coefs)
```
